Quantum H theorem, variational entropy, and relaxation processes

### 2020-10-09 notasparaarticulo.tex
classic limit was intruduced, conclusions and introduction were modified and the H functional was modified and references were added.

\section{Abstract}
We review in detail the quantum formulation of the Boltzamnns' H theorem. Starting from a H functional defined in the energy space. By means a variational procedure, we demonstrate the analogous of the quantum H theorem. At difference of the Tolmans' procedure to demonstrate the H theorem,  we start from the hypothesis of an out of equilibrium, spatially non homogeneous, system, and show that when $\frac{dH}{dt}=0$, the system reaches the equilibrium condition and then proposed H functional becomes the well known expression of the entropy of an ideal quantum gas. The time evolution of the system happens in such a way that $\frac{dH}{dt}<0$. Based on this, we propose a theoretical scheme that allows to describe the time evolution toward the equilibrium of an ideal quantum gas, it is the relaxation process of an out equilibrium.

%Variational entropy is defined as new $H$ functional that, performing varitational procedures, the Bose-Einstein and Fermi-Dirac distributions are obtained in the quantum scheme and the Maxwell-Boltzman distribution in the classical scheme. The $H$ theorem is proved in the quantum and classical schemes making use of the system is in a relaxation process. In t ssical scheme, a new $H$ functional is proposed according to the $H$ functional proposed by Boltzman. Besides, a ent expression for quantum systems is found in the varational procedure.%

@@

%revisar la redaccion y coherencia de estas ideas% 
One of those discussions is the propose of Tolman. He established the quantum $H$ theorem \cite{tolman} using the probability transition relation, the random phases hypothesis and a homogeneous distribution function. The lack of the quantum Boltzmann transport equation in analogy to classical statistical mechanics has stopped the development of the quantum transport formulation and the quantum non-equilibrium formulation. Another discission is the prove of the quantum $H$ theorem using n-colissions. This idea completes the $H$ theorem in a classical and quantum scheme \cite{binary}. There are some special conditions in quantum mechanics where the second law of thermodynamics is violated \cite{contradictions}, making that the quantum $H$ theorem is not thoroughly understood.\\
In transport phenomena, the Boltzmann transport equation is solved using the zero and first order approximation. When one uses the zero order approximation, one finds that the total heat flux in the system is zero, that is a contradiction. However, when one applies the first order approximation, one finds the evolution operator expressed by the left expression from the Boltzmann transport equation. That evolution operator tends to zero when time tends to infinity, this is the equilibrium condition \cite{reif}. This result, that describes inhomogeneties, is not expressed in the $H$ theorem.\\
In order to contribute to a consistent $H$ theorem, in this paper, the quantum and classical $H$ theorem are proved proposing a new $H$ functional for both cases considering a non-homogeneous distribution function and a system divided into imaginary cells. With those assumptions, particles transport is allowed inside the system. Performing variational procedures, one finds that, in the quantum case, Bose-Einstein and Fermi-Dirac distributions minimize the $H$ functional and, in the classical case, Maxwell-Boltzmann distribution minimize the other $H$ functional. A expression of entropy to quantum systems will be obtained in the variational procedure. In other hand, making use of the hypothesis that the system is in a relaxation process and the non-homogeneous distribution function, the $H$ theorem will be proved for both cases.

\section{The equilibrium of dilute gases: a review}
We recall that the classical Kinetic description of the equilibrium condition of a dilute gas as proposed by Boltzmann, can be done by means the Maxwell-Boltzmann distribution which is obtained imposing the condition dH/dt=0, where $H$ is defined as
\begin{equation}
    H=\int f(v,t) \ln f(v,t) \mathrm{d}^{3}v,
\end{equation} \label{hfunctional} 
where $f$ is the homogeneous distribution function.
In Quantum Mechanics the equilibrium of an ideal gas is done by means of the Fermi-Dirac or the Bose-Einstein distributions that can be obtained imposing the equiprobabilities and the random phases hypothesis and applying the quantum methods. The previous method in classical Kinetic description is not developed in quantum mechanics. A very short review for a classical and quantum mechanics about the H-theorem is presented in the next subsections.

\subsection{Classical approach}
In the classical version, the Boltzmann transport equation is used to shown the $H$-theorem. That equation is
\begin{equation}
    \left( \frac{\partial }{\partial t}+v_1 \cdot \nabla_r +\frac{F}{m} \cdot \nabla_{v_1} \right)f_1=\int \mathrm{d}\Omega \int \mathrm{d}^{3}v_2\sigma(\Omega)|v_1-v_2|(f_2'f_1'-f_2f_1),
\end{equation}
where $\Omega$ is the solid angle, $\sigma$ is the scattering section, $F$ the external force applied to the system and $f_1, f_2$ are the distribution function of the particles 1 and 2 respectively. Defining the functional (\ref{hfunctional}), the $H$ theorem is proved derivating the $H$ functional and applying the Boltzmann transport equation. The direction of the evolution of any system is predicted by the $H$-theorem. That direction always will be the equilibrium.
\
\subsection{Quantum approach}
In the quantum version presented by Tolman, the $H$ functional is defined as 
\begin{equation}
    H=\sum_i n_i \ln n_i -(n_i\pm g_i)\ln (g_i \pm n_i)\pm g_i\ln g_i, \label{quantumh}
\end{equation}
where $n_i$ is the occupation number and $g_i$ is a energy cell. Also, an equation for the rate of change in the number of particles in group $\kappa$ is obtained
\begin{eqnarray}
    \frac{d n_{\kappa}}{dt}&=&-\sum_{\lambda,(\mu \nu)}A_{\kappa\lambda,\mu\nu} n_{\kappa}n_{\lambda}(g_{\mu}\pm n_{\mu})(g_{\nu}\pm n_{\nu})\nonumber \\
    &&+\sum_{\lambda,(\mu \nu)}A_{\mu\nu,\kappa\lambda} n_{\mu}n_{\nu}(g_{\kappa}\pm n_{\kappa})(g_{\lambda}\pm n_{\lambda}),\label{changen}
\end{eqnarray}
where $A_{\kappa\lambda,\mu\nu}$ is a tensor expressed as
\begin{equation}
  A_{\kappa\lambda,\mu\nu}=\frac{4\pi^{2}}{h}\frac{|I_1\pm I_2|^2}{\Delta \epsilon},
\end{equation}
where $\epsilon$ is the energy and $|I_1-I_2|^2=|V_{mn,kl}|^2$ where $V_{mn,kl}$ is the perturbation matrix of binary particles. Those expression are obtained by the perturbation theory.\\
The $H$-theorem is proved using (\ref{changen}) into (\ref{quantumh}). This prove is in the same way as in the classical version.  That prove is developed taking account of the homogeneous distribution function, as in the classical version.
\section{Ensemble approach: quantum scheme}

We begin by defining the form of the functional $H$

@@

    H(t)&=&\sum_{M} \int_V \sum_{n} [ f_{Mn}(\epsilon_{n},t) \ln f_{Mn}(\epsilon_{n},t)\nonumber \\
    &&\pm (1 \mp f_{Mn}(\epsilon_{n},t)) \ln (1 \mp f_{Mn}(\epsilon_{n},t)) ]\mathrm{d}V \label{entropy}.

@@

global system in turn consists of a set of $K$ small cells, whose local distribution functions, $f_{Mn}(\epsilon_{n},t)$, are
identially zero outside the cell’s volume. The distribution function will depend on the position $\bar{r}$, the energy per cell $\epsilon_{n}$ and time $t$. We will use the uppercase latin index to denote cell numbers, and unless stated otherwise, $M = 1, 2 . . . , K$. Without loss of generality, we choose these cells to have

@@

    H_M&=& \int_V \sum_{n} [ f_{Mn}(\epsilon_{n},t) \ln f_{Mn}(\epsilon_{n},t)\nonumber \\
    &&\pm (1 \mp f_{Mn}(\epsilon_{n},t)) \ln (1 \mp f_{Mn}(\epsilon_{n},t)) ]\mathrm{d}V \label{entropycell},\\
    {\mathcal{N}}_M&=&\int_V \sum_{n}f_{Mn}(\epsilon_{n} ,t)\mathrm{d}V, \nonumber \\
{\mathcal{E}}_M&=&\int_V \sum_{n}f_{Mn}(\epsilon_{n},t)\epsilon_{n}\mathrm{d}V.

@@

    &&\sum_{M} \left[\int_V \sum_{n}f_{Mn}(\epsilon_{n} ,t)\mathrm{d}V\right]=\sum_{M} {\mathcal{N}}_{M}=N; \nonumber \\
    &&\sum_{M}\left[\int_V \sum_{n}f_{Mn}(\epsilon_{n},t)\epsilon_{Mn}\mathrm{d}V\right]=\sum_M {\mathcal{E}}_M=E, \label{restriccions1}

@@

where $\alpha_M(t), \beta_M(t)$ are the Lagrange multipliers. \textcolor{red}{We can obtain from the variational procedure that 

@@

\end{equation}}

@@

      H(t)&=&\sum_M \sum_n \int_V \left[\left(\frac{1}{e^{-\bar{\alpha}-\bar{\beta}\epsilon_{n}}\pm 1} \right)\ln \left(\frac{1}{e^{-\bar{\alpha}-\bar{\beta}\epsilon_{n}}} \right) \right]\mathrm{d}V\nonumber \\
      &&\pm \int_V \ln \left[\prod_{M} \prod_{n}\left(1 \mp \frac{1}{e^{-\bar{\alpha}-\bar{\beta}\epsilon_{n}}\pm 1} \right) \right]\mathrm{d}V\label{H-entropy}.

@@

\begin{equation}
   H(t)=\int_{}^{}\int_{}^{} f(r,v,t) \ln f(r,v,t)\mathrm{d}^3v \mathrm{d}^3r \label{CH2}.
\end{equation}{}
The functional (\ref{CH2}) contains a no-homogeneous distribution function but keeps the same form of the Boltzman's $H$ functional. Excluding the homogeneous distribution function hypothesis, one uses the same procedure, that it was shown in the previous section, to prove the $H$-theorem for relaxation processes.\\ 
If one considers the microcanonical ensemble conditions 
\begin{equation}
    \int_{}^{}\int_{}^{}f(r,v,t)\mathrm{d}^3v \mathrm{d}^3r=N, \ \ \ \int_{}^{}\int_{}^{}f(r,v,t)\epsilon(v)\mathrm{d}^3v \mathrm{d}^3r=E \label{micro},
\end{equation}{}
one obtains the distribution function that minimizes $H$
\begin{eqnarray}
\frac{\delta H}{\delta f(r',v')}&=&\int_{}^{}\frac{\delta}{\delta f(r',v')}\left[f(r,v)\ln f(r,v)  \right]\mathrm{d}^3v \mathrm{d}^3r-\int_{}^{}\alpha(r)\int_{}^{}\frac{\delta f(r,v)}{\delta f(r',v')}\mathrm{d}^3v \mathrm{d}^3r\nonumber \\
&&-\int_{}^{}\beta(r) \int_{}^{}\epsilon(v)\frac{\delta f(r,v)}{\delta f(r',v')}\mathrm{d}^3v \mathrm{d}^3r\nonumber \\
&=&\ln f(r',v')+1-\alpha(r')-\beta(r') \epsilon(v')=0.
\end{eqnarray}{}
isolating the distribution function 
\begin{eqnarray}
\ln f(r',v')&=&\alpha(r')+\beta(r') \epsilon(v')-1 \ \ \  \Rightarrow \ \ \ f(r',v')=e^{\alpha(r') +\beta(r') \epsilon(v')-1} \nonumber \\
&=&Ce^{\alpha(r')+\beta(r') \epsilon(v') } \label{relacion1},
\end{eqnarray}{}
where $C$ is a constant. It is trivial that using the variational procedures, one obtains the form of the Maxwell-Boltzmann distribution. \\
As $H$ in equilibrium is a density of entropy, it is trivial to think that Lagrange multipliers do not depend on the position because of the system is in equilibrium and consequently, the distribution function is homogeneous, that is
\begin{equation}
    f(v)=Ce^{\alpha+\beta \epsilon(v)}.
\end{equation}{} 
The constant $C$ can be omitted defining the following $H$ functional
\begin{equation}
   H(t)=\int_{}^{}\int_{}^{} \left[f(r,v,t) \ln f(r,v,t)-f(r,v,t)\right]\mathrm{d}^3v \mathrm{d}^3r \label{CH3},
\end{equation}{}
  but $\int \int f(r,v,t)\mathrm{d}^3v \mathrm{d}^3r=N$ where $N$ is the total particle number. As $N$ is a constant and the functional $H$ has sense when one calculates the change of $H$, this constant can be omitted. 
\section{H for relaxation processes in out of equilibrium systems}

@@

The functional H has the following form
\begin{equation}
    H(t)=\sum_{M} \int_V\sum_{n} \left[ f_{Mn} \ln f_{Mn} \pm (1 \mp f_{Mn}) \ln (1 \mp f_{Mn}) \right]\mathrm{d}V \label{entropy2},
\end{equation}
where, for simplicity, we have defined $f(\bar{r},\epsilon_{n},t)\equiv f_{Mn}$.
These functional has the following microcanonical restrictions
\begin{eqnarray}
        \int_V\sum_{n}f_{Mn}\mathrm{d}V=\bar{\mathcal{N}}+\Delta_M(t); \ \ \ \ \int_V\sum_{n}\epsilon_{n}f_{Mn}\mathrm{d}V=\bar{\mathcal{E}}+ \delta_M(t), \label{restrictionoutside}
  \end{eqnarray}
  where $\bar {\mathcal{N}}$ and $\bar{\mathcal{E}}$ are the local particle number and the local energy in equilibrium
  \begin{equation}
      \bar{\mathcal{N}}=\int_V \sum_n \bar{f}_n; \ \ \ \ \bar{\mathcal{E}}=\int_V \sum_n \epsilon_n\bar{f}_n,
  \end{equation}
  and $\Delta_M,\delta_M$ could be seen as a deviation from $\bar{\mathcal{N}}$ and $\bar{\mathcal{E}}$ respectively, with $\Delta_M(t)\ll \bar{\mathcal{N}}$ and $\delta_M(t) \ll \bar{\mathcal{E}}$. Those condition described a system out of equilibrium but no so far from it.  \\
If one performs the derivative of the functional H with respect to time, we obtain
\begin{equation}
   \frac{dH(t)}{dt}= \sum_n \sum_M \int_{V}\dot{f}_{nM}(t)\ln \left[ \frac{f_{nM}(t)}{1\mp f_{nM}(t)} \right]\mathrm{d}V.\label{deltaH}
\end{equation}{}
Using the first-order approximation
\begin{eqnarray}
   f_{Mn}=\bar{f}_{n}(1+g_{Mn}): \ \ \ \bar{f}_{n}\gg \bar{f}_{n}|g_{Mn}|, \label{firstorder}
\end{eqnarray}{}
we get

@@

    \frac{dH(t)}{dt}&=&\sum_n \sum_M \int_V\bar{f}_{n}\ln \left[ \frac{\bar{f}_{n}(1+g_{nM})}{1\mp \bar{f}_{n} (1+ g_{nM})} \right]\dot{g}_{nM}\mathrm{d}V \nonumber \\
    &=&\sum_n \sum_M \int_V\bar{f}_n \left \{ \ln [\bar{f}_n+\bar{f}_n g_{nM}]\dot{g}_{nM}-\ln [1\mp\bar{f}_n\mp\bar{f}_n g_{nM}]\dot{g}_{nM}  \right \}\mathrm{d}V.\nonumber \\

@@

    \frac{dH}{dt}&=&\sum_n \sum_M \int_V\bar{f}_n\left \{ (\ln \bar{f}_n+ g_{nM})\dot{g}_{nM}\right\}\mathrm{d}V \nonumber \\
    &&-\sum_{n}\sum_{M}\int_V\bar f_n\left\{ \left( \ln[1\mp\bar{f}_n]\mp \left[\frac{\bar{f}_n}{1\mp\bar{f}_n} \right] g_{nM}\right)\dot{g}_{nM} \right \}\mathrm{d}V.\label{cambioH2}

@@

    \frac{dH}{dt}&=&\sum_n \sum_M \int_V\bar{f}_n\left \{ (\bar{\alpha}+\bar{\beta}{\epsilon}_n)\dot{g}_{nM}+ g_{nM}\left(1\pm e^{\bar{\alpha}+\bar{\beta}{\epsilon}_n}\right)\dot{g}_{nM} \right \}\mathrm{d}V. \nonumber \\

@@

On the other hand, one can obtain from the restrictions the following expressions

@@

    &&\int_V\sum_n \bar{f}_n g_{nM}\mathrm{d}V=\Delta_M(t) \ \  \Rightarrow \ \ \int_V \sum_n \bar{f}_n \dot{g}_{nM}\mathrm{d}V=\dot{\Delta}_M(t), \nonumber \\
    &&\sum_n \int_V \bar{f}_n g_{nM}\epsilon_n\mathrm{d}V=\delta_M(t) \ \  \Rightarrow \ \ \int_V \sum_n \bar{f}_n \dot{g}_{nM}\epsilon_n\mathrm{d}V=\dot{\delta}_M(t)

@@

and as a consequence of $\sum_M \Delta_M(t)=\sum_M \delta_M(t)=0$, one finds

@@

    \sum_M \dot{\Delta}_M(t)=\sum_M \dot{\delta}_{M}(t)=0.

@@

Substituting the previous expression to (\ref{cambioH3}) one obtains

@@

   \frac{dH}{dt}&=&  \sum_n e^{\bar{\alpha}+\bar{\beta}\epsilon_n}\sum_M \int_V g_{nM}\dot{g}_{nM}\mathrm{d}V. \nonumber \\ \label{cambioH4}

@@

    \frac{dH}{dt}=\sum_n  e^{\bar{\alpha}+\bar{\beta}\epsilon_n}\left(\sum_J ^{L}\int_V g^{+}_{nJ}\dot{g}^{+}_{nJ}\mathrm{d}V+\sum^{N}_J \int_V g^{-}_{nJ}\dot{g}^{-}_{nJ}\mathrm{d}V \right), \label{cambioH5}

@@

    \frac{dH}{dt}=-\sum_n  e^{\bar{\alpha}+\bar{\beta}\epsilon_n}\left(\sum_J ^{L}\int_V |g^{+}_{nJ}||\dot{g}^{+}_{nJ}|\mathrm{d}V+\sum^{N}_J \int_V |g^{-}_{nJ}||\dot{g}^{-}_{nJ}|\mathrm{d}V \right), \label{cambioH6}

>>>>

\section{Abstract}
We review in detail the quantum formulation of the Boltzamnn's H theorem. Starting from a H functional defined in the energy space. By means a variational procedure, we demonstrate the analogous of the quantum H theorem. At difference of the Tolman's procedure to demonstrate the H theorem,  we start from the hypothesis of an out of equilibrium, spatially non homogeneous system, and show that when $\frac{dH}{dt}=0$, the system reaches the equilibrium condition and then proposed H functional becomes the well known expression of the entropy of an ideal quantum gas. The time evolution of the system happens in such a way that $\frac{dH}{dt}<0$. Based on this, we propose a theoretical scheme that allows to describe the time evolution toward the equilibrium of an ideal quantum gas, it is the relaxation process of an out equilibrium, Finally, in the limit of high energy levels, the Boltzmann's $H$ functional is recovered.

@@

%revisar la redaccion y coherencia de estas ideas% 
%One of those discussions is the propose of Tolman. He established the quantum $H$ theorem \cite{tolman} using the probability transition relation, the random phases hypothesis and a homogeneous distribution function. The lack of the quantum Boltzmann transport equation in analogy to classical statistical mechanics has stopped the development of the quantum transport formulation and the quantum non-equilibrium formulation. Another discission is the prove of the quantum $H$ theorem using n-colissions. This idea completes the $H$ theorem in a classical and quantum scheme \cite{binary}. There are some special conditions in quantum mechanics where the second law of thermodynamics is violated \cite{contradictions}, making that the quantum $H$ theorem is not thoroughly understood.\\
%In transport phenomena, the Boltzmann transport equation is solved using the zero and first order approximation. When one uses the zero order approximation, one finds that the total heat flux in the system is zero, that is a contradiction. However, when one applies the first order approximation, one finds the evolution operator expressed by the left expression from the Boltzmann transport equation. That evolution operator tends to zero when time tends to infinity, this is the equilibrium condition \cite{reif}. This result, that describes inhomogeneties, is not expressed in the $H$ theorem.\\
%In order to contribute to a consistent $H$ theorem, in this paper, the quantum and classical $H$ theorem are proved proposing a new $H$ functional for both cases considering a non-homogeneous distribution function and a system divided into imaginary cells. With those assumptions, particles transport is allowed inside the system. Performing variational procedures, one finds that, in the quantum case, Bose-Einstein and Fermi-Dirac distributions minimize the $H$ functional and, in the classical case, Maxwell-Boltzmann distribution minimize the other $H$ functional. A expression of entropy to quantum systems will be obtained in the variational procedure. In other hand, making use of the hypothesis that the system is in a relaxation process and the non-homogeneous distribution function, the $H$ theorem will be proved for both cases.

\section{The equilibrium of dilute gases: a review}
\subsection{Classical approach}
In the classical version, the Boltzmann transport equation is used to shown the $H$-theorem. That equation is
\begin{equation}
    \left( \frac{\partial }{\partial t}+v_1 \cdot \nabla_r +\frac{F}{m} \cdot \nabla_{v_1} \right)f_1=\int \mathrm{d}\Omega \int \mathrm{d}^{3}v_2\sigma(\Omega)|v_1-v_2|(f_2'f_1'-f_2f_1),
\end{equation}
where $\Omega$ is the solid angle, $\sigma$ is the scattering section, $F$ the external force applied to the system and $f_1, f_2$ are the distribution function of the particles 1 and 2 respectively.\\
The $H$ theorem can be proved using the $H$ functional defined by Boltzmann 
\begin{equation}
    H=\int f(v,t) \ln f(v,t) \mathrm{d}^{3}v,
\end{equation} \label{hfunctional}
where $f(v,t)$ is a homogeneous distribution function. The $H$ theorem establishes that if $f(v,t)$ satisfies the Boltzmann transport equation, then $\frac{dH}{dt}\leq 0$. The system arrives to the equilibrium state when $\frac{dH}{dt}=0$ and due to this, the distribution function to the system in equilibrium corresponds to the Maxwell-Boltzmann distribution
\subsection{Quantum approach}
The $H$ functional in quantum mechanics presented by Tolman \cite{tolman} is defined as 
\begin{equation}
    H=\sum_i n_i \ln n_i -(n_i\pm g_i)\ln (g_i \pm n_i)\pm g_i\ln g_i, \label{quantumh}
\end{equation}
where $n_i$ is the occupation number and $g_i$ is a energy cell. Also, an equation for the rate of change in the number of particles in group $\kappa$ is obtained
\begin{eqnarray}
    \frac{d n_{\kappa}}{dt}&=&-\sum_{\lambda,(\mu \nu)}A_{\kappa\lambda,\mu\nu} n_{\kappa}n_{\lambda}(g_{\mu}\pm n_{\mu})(g_{\nu}\pm n_{\nu})\nonumber \\
    &&+\sum_{\lambda,(\mu \nu)}A_{\mu\nu,\kappa\lambda} n_{\mu}n_{\nu}(g_{\kappa}\pm n_{\kappa})(g_{\lambda}\pm n_{\lambda}),\label{changen}
\end{eqnarray}
where $A_{\kappa\lambda,\mu\nu}$ is a tensor expressed as
\begin{equation}
  A_{\kappa\lambda,\mu\nu}=\frac{4\pi^{2}}{h}\frac{|I_1\pm I_2|^2}{\Delta \epsilon},
\end{equation}
where $\epsilon$ is the energy and $|I_1-I_2|^2=|V_{mn,kl}|^2$ where $V_{mn,kl}$ is the perturbation matrix of binary particles. Those expression are obtained by the perturbation theory.\\
The quantum $H$-theorem is proved using (\ref{changen}) into (\ref{quantumh}). This prove is in the same way as in the classical version.  That prove is developed taking account of the homogeneous distribution function, as in the classical version.\\
In addition to this, Tolman showed that (\ref{quantumh}) can be reduced in the limit of high energy levels to
\begin{equation}
    H = \sum_{\kappa} (n_{\kappa} \ln n_{\kappa} - n_{\kappa} \ln g_{\kappa}).
\end{equation}
This expression can be obtained also from the Boltzmann $H$ functional by means defining 
\begin{equation}
    f=\frac{n_{\kappa}}{ g_{\kappa}}.
\end{equation}
\section{Ensemble approach}
\subsection{Quantum scheme}

\textcolor{red}{We begin by defining the form of the functional $H$

@@

    H(t)&=&\sum_{M} \sum_{n} [ f_{Mn}(\epsilon_{n},t) \ln f_{Mn}(\epsilon_{n},t)\nonumber \\
    &&\pm (1 \mp f_{Mn}(\epsilon_{n},t)) \ln (1 \mp f_{Mn}(\epsilon_{n},t)) ] \delta V_M \delta \epsilon_n \label{entropy}.

@@

global system in turn consists of a set of $K$ small cells with a constant volume $V_M$, whose local homogeneous distribution functions is $f_{Mn}(\epsilon_{n},t)$. The distribution function will depend on the position of the cells, the energy per cell $\epsilon_{n}$ and time $t$ and the energy spectra will be divided into $g_n$ volumes. We will use the uppercase latin index to denote cell numbers, and unless stated otherwise, $M = 1, 2 . . . , K$. Without loss of generality, we choose these cells to have

@@

    H_M&=&  \sum_{n} [ f_{Mn}(\epsilon_{n},t) \ln f_{Mn}(\epsilon_{n},t)\nonumber \\
    &&\pm (1 \mp f_{Mn}(\epsilon_{n},t)) \ln (1 \mp f_{Mn}(\epsilon_{n},t)) ] \delta \epsilon_n \label{entropycell},\\
    {\mathcal{N}}_M&=& \sum_{n}f_{Mn}(\epsilon_{n} ,t) \delta \epsilon_n, \nonumber \\
{\mathcal{E}}_M&=& \sum_{n}f_{Mn}(\epsilon_{n},t)\epsilon_{n} \delta \epsilon_n.

@@

    &&\sum_{M} \left[ \sum_{n}f_{Mn}(\epsilon_{n} ,t)\right]\delta V_M \delta \epsilon_n=\sum_{M} {\mathcal{N}}_{M} \delta \epsilon_n=N; \nonumber \\
    &&\sum_{M}\left[ \sum_{n}f_{Mn}(\epsilon_{n},t)\epsilon_{Mn}\right]\delta V_M \delta g_n=\sum_M {\mathcal{E}}_M \delta \epsilon_n=E, \label{restriccions1}

@@

where $\alpha_M(t), \beta_M(t)$ are the Lagrange multipliers. We can obtain from the variational procedure that 

@@

@@

      H(t)&=&\sum_M \sum_n  \left[\left(\frac{1}{e^{-\bar{\alpha}-\bar{\beta}\epsilon_{n}}\pm 1} \right)\ln \left(\frac{1}{e^{-\bar{\alpha}-\bar{\beta}\epsilon_{n}}} \right) \right]\nonumber \\
      &&\pm  \ln \left[\prod_{M} \prod_{n}\left(1 \mp \frac{1}{e^{-\bar{\alpha}-\bar{\beta}\epsilon_{n}}\pm 1} \right) \right]\delta V_M \delta \epsilon_n\label{H-entropy}.

@@

\begin{equation}
   H(t)=\int_{}^{}\int_{}^{} f(r,v,t) \ln f(r,v,t)\mathrm{d}^3v \mathrm{d}^3r \label{CH2}.
\end{equation}{}
The functional (\ref{CH2}) contains a no-homogeneous distribution function but keeps the same form of the Boltzman's $H$ functional. Excluding the homogeneous distribution function hypothesis, one uses the same procedure, that it was shown in the previous section, to prove the $H$-theorem for relaxation processes.\\ 
If one considers the microcanonical ensemble conditions 
\begin{equation}
    \int_{}^{}\int_{}^{}f(r,v,t)\mathrm{d}^3v \mathrm{d}^3r=N, \ \ \ \int_{}^{}\int_{}^{}f(r,v,t)\epsilon(v)\mathrm{d}^3v \mathrm{d}^3r=E \label{micro},
\end{equation}{}
one obtains the distribution function that minimizes $H$
\begin{eqnarray}
\frac{\delta H}{\delta f(r',v')}&=&\int_{}^{}\frac{\delta}{\delta f(r',v')}\left[f(r,v)\ln f(r,v)  \right]\mathrm{d}^3v \mathrm{d}^3r-\int_{}^{}\alpha(r)\int_{}^{}\frac{\delta f(r,v)}{\delta f(r',v')}\mathrm{d}^3v \mathrm{d}^3r\nonumber \\
&&-\int_{}^{}\beta(r) \int_{}^{}\epsilon(v)\frac{\delta f(r,v)}{\delta f(r',v')}\mathrm{d}^3v \mathrm{d}^3r\nonumber \\
&=&\ln f(r',v')+1-\alpha(r')-\beta(r') \epsilon(v')=0.
\end{eqnarray}{}
isolating the distribution function 
\begin{eqnarray}
\ln f(r',v')&=&\alpha(r')+\beta(r') \epsilon(v')-1 \ \ \  \Rightarrow \ \ \ f(r',v')=e^{\alpha(r') +\beta(r') \epsilon(v')-1} \nonumber \\
&=&Ce^{\alpha(r')+\beta(r') \epsilon(v') } \label{relacion1},
\end{eqnarray}{}
where $C$ is a constant. It is trivial that using the variational procedures, one obtains the form of the Maxwell-Boltzmann distribution. \\
As $H$ in equilibrium is a density of entropy, it is trivial to think that Lagrange multipliers do not depend on the position because of the system is in equilibrium and consequently, the distribution function is homogeneous, that is
\begin{equation}
    f(v)=Ce^{\alpha+\beta \epsilon(v)}.
\end{equation}{} 
The constant $C$ can be omitted defining the following $H$ functional
\begin{equation}
   H(t)=\int_{}^{}\int_{}^{} \left[f(r,v,t) \ln f(r,v,t)-f(r,v,t)\right]\mathrm{d}^3v \mathrm{d}^3r \label{CH3},
\end{equation}{}
  but $\int \int f(r,v,t)\mathrm{d}^3v \mathrm{d}^3r=N$ where $N$ is the total particle number. As $N$ is a constant and the functional $H$ has sense when one calculates the change of $H$, this constant can be omitted. 
\section{$H$-functional for relaxation processes in out of equilibrium systems}

@@

The functional H \textcolor{red}{(variational entropy)} has the following form
\begin{equation}
    H(t)=\sum_{M} \sum_{n} \left[ f_{Mn} \ln f_{Mn} \pm (1 \mp f_{Mn}) \ln (1 \mp f_{Mn}) \right]\delta V_M \delta \epsilon_n \label{entropy2},
\end{equation}
where, for simplicity, we have defined $f(\bar{r},\epsilon_{n},t)\equiv f_{Mn}$.
These functional has the following microcanonical restrictions
\begin{eqnarray}
        \sum_{n}f_{Mn} \delta \epsilon_n=\bar{\mathcal{N}}+\Delta_M(t); \ \ \ \ \sum_{n}\epsilon_{n}f_{Mn} \delta \epsilon_n=\bar{\mathcal{E}}+ \delta_M(t), \label{restrictionoutside}
  \end{eqnarray}
  where $\bar {\mathcal{N}}$ and $\bar{\mathcal{E}}$ are the local particle number and the local energy in equilibrium
  \begin{equation}
      \bar{\mathcal{N}}= \sum_n \bar{f}_n \delta \epsilon_n; \ \ \ \ \bar{\mathcal{E}}= \sum_n \epsilon_n\bar{f}_n \delta \epsilon_n,
  \end{equation}
  and $\Delta_M,\delta_M$ could be seen as a deviation from $\bar{\mathcal{N}}$ and $\bar{\mathcal{E}}$ respectively, with $\Delta_M(t)\ll \bar{\mathcal{N}}$ and $\delta_M(t) \ll \bar{\mathcal{E}}$. Those condition described a system out of equilibrium but no so far from it.  \\
If one performs the derivative of the functional H with respect to time, we obtain
\begin{equation}
   \frac{dH(t)}{dt}= \sum_n \sum_M \dot{f}_{nM}(t)\ln \left[ \frac{f_{nM}(t)}{1\mp f_{nM}(t)} \right]\delta V_M \delta \epsilon_n.\label{deltaH}
\end{equation}{}
Using the first-order approximation
\begin{eqnarray}
   f_{Mn}=\bar{f}_{n}(1+g_{Mn}): \ \ \ \bar{f}_{n}\gg \bar{f}_{n}|g_{Mn}|, \label{firstorder}
\end{eqnarray}{}
we get

@@

    \frac{dH(t)}{dt}&=&\sum_n \sum_M \bar{f}_{n}\ln \left[ \frac{\bar{f}_{n}(1+g_{nM})}{1\mp \bar{f}_{n} (1+ g_{nM})} \right]\dot{g}_{nM}\delta V_M \delta \epsilon_n \nonumber \\
    &=&\sum_n \sum_M \bar{f}_n \left \{ \ln [\bar{f}_n+\bar{f}_n g_{nM}]\dot{g}_{nM}-\ln [1\mp\bar{f}_n\mp\bar{f}_n g_{nM}]\dot{g}_{nM}  \right \}\delta V_M \delta \epsilon_n.\nonumber \\

@@

    \frac{dH}{dt}&=&\sum_n \sum_M \bar{f}_n\left \{ (\ln \bar{f}_n+ g_{nM})\dot{g}_{nM}\right\}\delta V_M \delta \epsilon_n \nonumber \\
    &&-\sum_{n}\sum_{M}\bar f_n\left\{ \left( \ln[1\mp\bar{f}_n]\mp \left[\frac{\bar{f}_n}{1\mp\bar{f}_n} \right] g_{nM}\right)\dot{g}_{nM} \right \}\delta V_M \delta \epsilon_n.\label{cambioH2}

@@

    \frac{dH}{dt}&=&\sum_n \sum_M \bar{f}_n\left \{ (\bar{\alpha}+\bar{\beta}{\epsilon}_n)\dot{g}_{nM}+ g_{nM}\left(1\pm e^{\bar{\alpha}+\bar{\beta}{\epsilon}_n}\right)\dot{g}_{nM} \right \}\delta V_M \delta \epsilon_n. \nonumber \\

@@

On the other hand, we obtain from the restrictions the following expressions

@@

    &&\sum_n \bar{f}_n g_{nM} \delta \epsilon_n=\Delta_M(t) \ \  \Rightarrow \ \  \sum_n \bar{f}_n \dot{g}_{nM} \delta \epsilon_n=\dot{\Delta}_M(t), \nonumber \\
    &&\sum_n  \bar{f}_n g_{nM}\epsilon_n \delta \epsilon_n=\delta_M(t) \ \  \Rightarrow \ \  \sum_n \bar{f}_n \dot{g}_{nM}\epsilon_n \delta \epsilon_n=\dot{\delta}_M(t)

@@

and as a consequence of $\sum_M \Delta_M(t) \delta V_M =\sum_M \delta_M(t)\delta V_M =0$, one finds

@@

    \sum_M \dot{\Delta}_M(t)\delta V_M =\sum_M \dot{\delta}_{M}(t)\delta V_M =0.

@@

Substituting the previous expression to (\ref{cambioH3}) we obtain

@@

   \frac{dH}{dt}&=&  \sum_n e^{\bar{\alpha}+\bar{\beta}\epsilon_n}\sum_M  g_{nM}\dot{g}_{nM}\delta V_M \delta \epsilon_n. \nonumber \\ \label{cambioH4}

@@

    \frac{dH}{dt}=\sum_n  e^{\bar{\alpha}+\bar{\beta}\epsilon_n}\left(\sum_J ^{L} g^{+}_{nJ}\dot{g}^{+}_{nJ}\delta V_M \delta \epsilon_n+\sum^{N}_J  g^{-}_{nJ}\dot{g}^{-}_{nJ}\delta V_M \delta \epsilon_n \right), \label{cambioH5}

@@

    \frac{dH}{dt}=-\sum_n  e^{\bar{\alpha}+\bar{\beta}\epsilon_n}\left(\sum_J ^{L} |g^{+}_{nJ}||\dot{g}^{+}_{nJ}|\delta V_M \delta \epsilon_n+\sum^{N}_J  |g^{-}_{nJ}||\dot{g}^{-}_{nJ}|\delta V_M \delta \epsilon_n \right), \label{cambioH6}
